Trends from 30-year observations of downward solar irradiance in Thessaloniki, Greece

### 2023-12-16 MDPI_submission.Rmd
Commit 2023-12-16 13:09

<!--
Η εισαγωγή είναι εντάξει. Ίσως θα μπορούσες να βάλεις στη 2η
παράγραφο ακόμη 2-3 σειρές για τις τάσεις του αεροζόλ και των νεφών
από τις εργασίες και να καταλήξεις με κάποια στοιχεία για το αεροζόλ
στη Θεσσαλονίκη από εργασίες δικές μας. Για τα νέφη δεν έχω υπόψη μου
καμία εργασία αλλά ίσως θα μπορούσε να ειπωθεί κάτι χρησιμοποιώντας
παγκόσμιες εκτιμήσεις για την περιοχή της Μεσογείου.
π.χ. https://link.springer.com/article/10.1007/s00382-022-06438-3
-->

@@

<!-- c) Clear sky identification -->
In order to be able to estimate the effect of the sky conditions on the long term variability of SDR, we created three datasets, by characterizing each one-minute measurement with a corresponding sky condition (i.e., all-sky, clear-sky and cloudy-sky).
To identify the clear-sky conditions we used a method proposed by
@Long2000
and by
@Reno2016,
which were adapted and configured for the site, as the authors suggest.

<!-- CS subjectivity -->
We have to note, that the definition of clear or cloudy sky, has some subjectivity, in any method of characterization.
As a result, the details of the definition are site specific, it relies on a combination of thresholds and comparisons with ideal radiation models and statistical analysis of different signal metrics.
The CSid algorithm was calibrated with the main focus, to identify the presence of clouds on the sky.
Despite the fine-tuning of the procedure, a few marginal cases exist, that have been identified manually as false positive or false negative but cannot affect the final results of the study.

@@

For completeness, we will provide below a brief overview of the clear sky identification algorithm (CSid), along with the site specific thresholds.
To calculate the reference clear sky $\text{SDR}_\text{CSref}$ we used the $\text{SDR}_\text{Haurwitz}$ derived by the radiation model of @Haurwitz1945, adjusted for our site with a factor $a$ (Eq.\ \ref{eq:ahau}), resulted by an iterative optimization process, as described by @Long2000 and @Reno2016.
The target of the optimization was the minimization of a function $f(a)$ (Eq.\ \ref{eq:minf}) and was accomplished with the algorithmic function "optimise", which is an implementation based on the work of @Brent1973, from the library "stats" of the R programming language [@RCT2023].
<!-- cost function -->
\begin{equation}
f(a) = \frac{1}{n}\sum_{i=1}^{n} ( \text{SDR}_{\text{CSid},i} - a \times \text{SDR}_{\text{testCSref},i} )^2 \label{eq:minf}
\end{equation}
where: $n$ is the total number of daylight records, $\text{SDR}_{\text{CSid},i}$ are the records identified as clear sky by CSid, $a$ is a hypothetical adjustment factor, and $\text{SDR}_{\text{testCSref},i}$ is any of the tested clear sky radiation models.

The optimization and the selection of the clear sky reference model, was performed on SDR observations for the period 2016 - 2021.
During the optimization, eight simple clear sky radiation models were tested (namely, Daneshyar-Paltridge-Proctor, Kasten-Czeplak, Haurwitz, Berger-Duffie, Adnot-Bourges-Campana-Gicquel, Robledo-Soler, Kasten and Ineichen-Perez), with a wide range of factors.
These models are described in more details by @Reno2012 and evaluated by @Reno2016.
We found, that Haurwitz's model, adjusted with the factor $a = `r signif(alpha,3)`$ yields one of the lowest root mean squared errors (RMSE), while the procedure, manages to characterize the majority of the data.
Thus, our clear sky reference is derived by the Eq.\ \ref{eq:ahau}.
<!-- adjusted Haurwitz -->
\begin{equation}
\text{SDR}_\text{CSref} = a \times \text{SDR}_\text{Haurwitz} = `r signif(alpha,3)` \times 1098 \times \cos(\theta) \times \exp \left( \frac{ - 0.057}{\cos(\theta)} \right) \label{eq:ahau}
\end{equation}
where: $\text{SDR}_\text{CSref}$ is the reference clear sky SDR, in $\text{W}\,\text{m}^{-2}$ and $\theta$ is the solar zenith angle (SZA).

The criteria that were used to identify whether a measurement was taken under clear-sky conditions are presented below.
A data point is flagged as "clear-sky" if all criteria are satisfied, otherwise it is considered to be "cloud-sky".
Each criterion was applied for a running window of $`r nt`$ consecutive one-minute measurements, and the characterization is assigned to the central value of the window.
Each parameter, was calculated both from the observations and the reference clear sky model, for each comparison.
The allowable range of variation is defined by the model-derived value of the parameter multiplied by a factor plus an offset.
The factors and the offsets were determined empirically, by manually inspecting each filter's performance on selected days and adjusting them accordingly during an iterative process.
The criteria are:
<!-- 1. MeanVIP -->
<!-- ### Mean value of irradiance during the time period. -->
<!--a) Mean of the measured $\overline{\text{SDR}}_i$ = \text{mean}_{n=i-5}^{n=i+5}[\text{SDR}_{n}]$ and running mean of the reference $\overline{\text{SDR}}_{\text{CSref},i}$ = \text{mean}_{n=i-5}^{n=i+5}[\text{SDR}_{\text{CSref},n}]$ (Eq. \ref{eq:MeanVIP}).-->

@@

<!-- ### Data selection -->
Due to the significant measurement uncertainty when the Sun is near the horizon, we have excluded all measurements with SZA greater than $`r 90-MIN_ELEVA`^\circ$.
Moreover, due to some obstructions around the site (hills and buildings), we excluded data with azimuth angle between
$`r FIBais_Az_1`^\circ$ and $`r FIBais_Az_2`^\circ$ with SZA greater than $`r 90 -  FIBais_Elev`^\circ$.
On the latter instances, Sun is systematically not visible from the instrument's location.
<!-- 1au -->
To make the measurements comparable throughout the dataset, we adjusted all one-minute radiometric values to the mean Sun - Earth distance.
<!-- TSI -->
Subsequently, we made all measurements relative to the Total Solar Irradiance (TSI) at $1\,\text{au}$, in order to compensate for the Sun's intensity variability, using a time series of satellite TSI observations.
The TSI data we used are part of the "NOAA Climate Data Record of Total Solar Irradiance" dataset [@Coddington2005].
The initial daily values of this dataset were interpolated to match the time step of our measurements.
In the final dataset
<!-- $`r SDR_N_All`$ -->
$`r signif(100 * SDR_N_Clear / SDR_N_All, 3)`\%$
of the 1-minute data were identified as under clear-sky conditions and
$`r signif(100 * SDR_N_Cloud / SDR_N_All, 3)`\%$
as under cloud-sky conditions.
<!-- The TSI data we use, is a combination of data from NOAA [@Coddington2005] (for `r as.Date(TSIinfo[TSI_Source == "NOAA", Start])` - `r as.Date(TSIinfo[TSI_Source == "NOAA", End])`) and adjusted data from @LASP2023 (for `r as.Date(TSIinfo[TSI_Source != "NOAA", Start])` - `r as.Date(TSIinfo[TSI_Source != "NOAA", End])`).-->

<!-- aggregation -->
In order to investigate the SDR trends, we implemented an appropriate aggregation scheme to the 1-minute data to derive a series in coarser time-scales.
To preserve the representativeness of the data we used the following criteria:
<!-- a) for daily mean values we exclude days with less than `r Daily_aggregation_N_lim` valid data points, -->
a) we accept only days with more than `r 100 * All_daily_ratio_lim`% of the daytime measurements present and valid,
b) on the dataset of clear- and cloudy-skies, we included only days with more than
`r 100 * Clear_daily_ratio_lim`% of the daytime measurements identified as clear or cloudy respectively,
c) monthly values were computed from daily means
only when at least `r Monthly_aggegation_N_lim` days were available.
To create the daily and monthly climatological means, we averaged the data based on the day of year and calendar month, respectively.
<!-- Seasonal -->
For the seasonal means we averaged the mean daily values in each season (Winter: December - February, Spring: March - May, etc.).
<!-- Deseasonalization -->
Finally, each data set was deseasonalized by subtracting the corresponding climatological annual cycle (daily or monthly) from the actual data.
To estimate the SZA effect on the SDR trends, the one-minute data were aggregated in $1^\circ$ SZA bins, separately for the morning and afternoon hours, and then were deseasonalized as mentioned above.

@@

```{r trendALL, echo=F, out.width='.70\\linewidth', fig.cap=paste0("Anomalies (\\%) of the daily all-sky SDR, relative to climatological values for ", year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)), ". The black line shows the long term linear trend.")}
knitr::include_graphics("./images/LongtermTrends-2.png")
```

>>>>

@@

<!-- ### Data selection -->
Due to the significant measurement uncertainty when the Sun is near the horizon, we have excluded all measurements with SZA greater than $`r 90-MIN_ELEVA`^\circ$.
Moreover, due to obstructions around the site (hills and buildings) which block the direct irradiance, we excluded data with azimuth angle in the range
$`r FIBais_Az_1`^{\circ}$ - $`r FIBais_Az_2`^{\circ}$ and with SZA greater than $`r 90 -  FIBais_Elev`^{\circ}$.
<!-- 1au -->
To make the measurements comparable throughout the dataset, we adjusted all one-minute data to the mean Sun - Earth distance.
<!-- TSI -->
Subsequently, we adjusted all measurements to the Total Solar Irradiance (TSI) at $1\,\text{au}$, in order to compensate for the Sun's intensity variability, using a time series of satellite TSI observations.
The TSI data we used are part of the "NOAA Climate Data Record of Total Solar Irradiance" dataset [@Coddington2005].
The initial daily values of this dataset were interpolated to match the time step of our measurements.

<!-- c) Clear sky identification -->
In order to estimate the effect of the sky conditions on the long term variability of SDR, we created three datasets by characterizing each one-minute measurement with a corresponding sky-condition flag (i.e., all-sky, clear-sky and cloudy-sky).
To identify the clear-cases we used a method proposed by
@Long2000
and by
@Reno2016,
which were adapted and configured for the site of Thessaloniki.
<!-- CS subjectivity -->
We note that all methods have some subjectivity in the definition of clear or cloudy sky cases.
As a result, the details of the definition are site specific and they rely on a combination of thresholds and comparisons with ideal radiation models and statistical analysis of different signal metrics.
The CSid algorithm was calibrated with the main focus, to identify the presence of clouds. Despite the fine-tuning
of the procedure, in a few marginal cases false positive or false
negative results were identified by manual inspection. However, due to
their small number, they cannot affect the final results of the study.
For completeness, we provide below a brief overview of the clear-sky
identification algorithm (CSid), along with the site specific
thresholds.

@@

To calculate the reference clear-sky $\text{SDR}_\text{CSref}$ we used the $\text{SDR}_\text{Haurwitz}$ derived by the radiation model of @Haurwitz1945, adjusted for our site with a factor $a$ (Eq.\ \ref{eq:ahau}), estimated through an iterative optimization process, as described by @Long2000 and @Reno2016.
The target of the optimization was the minimization of a function $f(a)$ (Eq.\ \ref{eq:minf}) and was accomplished with the algorithmic function "optimise", which is an implementation based on the work of @Brent1973, from the library "stats" of the R programming language [@RCT2023].
<!-- cost function -->
\begin{equation}
f(a) = \frac{1}{n}\sum_{i=1}^{n} ( \text{SDR}_{\text{CSid},i} - a \times \text{SDR}_{\text{testCSref},i} )^2 \label{eq:minf}
\end{equation}
where: $n$ is the total number of daylight data, $\text{SDR}_{\text{CSid},i}$ are the data identified as clear sky by CSid, $a$ is a site-specific adjustment factor, and $\text{SDR}_{\text{testCSref},i}$ is the SDR derived by any of the
tested clear-sky radiation models.

The optimization and the selection of the clear sky reference model, was performed on SDR observations for the period 2016 - 2021.
During the optimization, eight simple clear sky radiation models were tested (namely, Daneshyar-Paltridge-Proctor, Kasten-Czeplak, Haurwitz, Berger-Duffie, Adnot-Bourges-Campana-Gicquel, Robledo-Soler, Kasten and Ineichen-Perez), with a wide range of factors.
These models are described in more details by @Reno2012 and evaluated by @Reno2016.
We found, that Haurwitz's model, adjusted with the factor $a = `r signif(alpha,3)`$ yields one of the lowest root mean squared errors (RMSE), while the procedure manages to characterize successfully the majority of the data.
Thus, our clear sky reference is derived by the Eq.\ \ref{eq:ahau}.
<!-- adjusted Haurwitz -->
\begin{equation}
\text{SDR}_\text{CSref} = a \times \text{SDR}_\text{Haurwitz} = `r signif(alpha,3)` \times 1098 \times \cos(\theta) \times \exp \left( \frac{ - 0.057}{\cos(\theta)} \right) \label{eq:ahau}
\end{equation}
where $\theta$ is the solar zenith angle (SZA).

The criteria that were used to identify whether a measurement was taken under clear-sky conditions are presented below.
A data point is flagged as "clear-sky" if all criteria are satisfied; otherwise it is considered as "cloud-sky".
Each criterion was applied for a running window of $`r nt`$ consecutive one-minute measurements, and the characterization was assigned to the central datum of the window.
Each parameter, was calculated both from the observations and the reference clear sky model, for each comparison.
The allowable range of variation is defined by the model-derived value of the parameter multiplied by a factor plus an offset.
The factors and the offsets were determined empirically, by manually inspecting each filter's performance on selected days and adjusting them accordingly during an iterative process.
The criteria are
listed below, together with the range of values within which the
respective parameter should fall in order to raise the clear-sky flag:
<!-- 1. MeanVIP -->
<!-- ### Mean value of irradiance during the time period. -->
<!--a) Mean of the measured $\overline{\text{SDR}}_i$ = \text{mean}_{n=i-5}^{n=i+5}[\text{SDR}_{n}]$ and running mean of the reference $\overline{\text{SDR}}_{\text{CSref},i}$ = \text{mean}_{n=i-5}^{n=i+5}[\text{SDR}_{\text{CSref},n}]$ (Eq. \ref{eq:MeanVIP}).-->

@@

In the final dataset
<!-- $`r SDR_N_All`$ -->
$`r signif(100 * SDR_N_Clear / SDR_N_All, 3)`\%$
of the 1-minute data were identified as under clear-sky conditions and
$`r signif(100 * SDR_N_Cloud / SDR_N_All, 3)`\%$
as under cloud-sky conditions.
<!-- The TSI data we use, is a combination of data from NOAA [@Coddington2005] (for `r as.Date(TSIinfo[TSI_Source == "NOAA", Start])` - `r as.Date(TSIinfo[TSI_Source == "NOAA", End])`) and adjusted data from @LASP2023 (for `r as.Date(TSIinfo[TSI_Source != "NOAA", Start])` - `r as.Date(TSIinfo[TSI_Source != "NOAA", End])`).-->

## Aggregation of data and statistical approach

In order to investigate the SDR trends which are the main focus of the
study, we implemented an appropriate aggregation scheme to the
one-minute data to derive a series in coarser time-scales.
To preserve the representativeness of the data we used the following criteria:
<!-- a) for daily mean values we exclude days with less than `r Daily_aggregation_N_lim` valid data points, -->
a) we accept only days with more than `r 100 * All_daily_ratio_lim`% of the daytime measurements present and valid,
b) on the dataset of clear- and cloudy-skies, we included only days with more than
`r 100 * Clear_daily_ratio_lim`% of the daytime measurements identified as clear or cloudy respectively,
c) monthly values were computed from daily means
only when at least `r Monthly_aggegation_N_lim` days were available.
To create the daily and monthly climatological means, we averaged the data based on the day of year and calendar month, respectively.
<!-- Seasonal -->
For the seasonal means we averaged the mean daily values in each season (Winter: December - February, Spring: March - May, etc.).
<!-- Deseasonalization -->
Finally, each data set was deseasonalized by subtracting the corresponding climatological annual cycle (daily or monthly) from the actual data.
To estimate the SZA effect on the SDR trends, the one-minute data were aggregated in $1^\circ$ SZA bins, separately for the morning and afternoon hours, and then were deseasonalized as mentioned above.

@@

```{r trendALL, echo=F, out.width='.70\\linewidth', fig.cap=paste0("Anomalies (\\%) of the daily all-sky SDR from the climatological mean for the period", year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)), ". The black line is the long term linear trend.")}
knitr::include_graphics("./images/LongtermTrends-2.png")
```

### 2023-12-16 MDPI_submission.tex
Commit 2023-12-16 13:09

In order to be able to estimate the effect of the sky conditions on the
long term variability of SDR, we created three datasets, by
characterizing each one-minute measurement with a corresponding sky
condition (i.e., all-sky, clear-sky and cloudy-sky). To identify the
clear-sky conditions we used a method proposed by \citet{Long2000} and
by \citet{Reno2016}, which were adapted and configured for the site, as
the authors suggest.

We have to note, that the definition of clear or cloudy sky, has some
subjectivity, in any method of characterization. As a result, the
details of the definition are site specific, it relies on a combination
of thresholds and comparisons with ideal radiation models and
statistical analysis of different signal metrics. The CSid algorithm was
calibrated with the main focus, to identify the presence of clouds on
the sky. Despite the fine-tuning of the procedure, a few marginal cases
exist, that have been identified manually as false positive or false
negative but cannot affect the final results of the study.

@@

For completeness, we will provide below a brief overview of the clear
sky identification algorithm (CSid), along with the site specific
thresholds. To calculate the reference clear sky
\(\text{SDR}_\text{CSref}\) we used the \(\text{SDR}_\text{Haurwitz}\)
derived by the radiation model of \citet{Haurwitz1945}, adjusted for our
site with a factor \(a\) (Eq.~\ref{eq:ahau}), resulted by an iterative
optimization process, as described by \citet{Long2000} and
\citet{Reno2016}. The target of the optimization was the minimization of
a function \(f(a)\) (Eq.~\ref{eq:minf}) and was accomplished with the
algorithmic function ``optimise'', which is an implementation based on
the work of \citet{Brent1973}, from the library ``stats'' of the R
programming language \citep{RCT2023}. \begin{equation}
f(a) = \frac{1}{n}\sum_{i=1}^{n} ( \text{SDR}_{\text{CSid},i} - a \times \text{SDR}_{\text{testCSref},i} )^2 \label{eq:minf}
\end{equation} where: \(n\) is the total number of daylight records,
\(\text{SDR}_{\text{CSid},i}\) are the records identified as clear sky
by CSid, \(a\) is a hypothetical adjustment factor, and
\(\text{SDR}_{\text{testCSref},i}\) is any of the tested clear sky
radiation models.

The optimization and the selection of the clear sky reference model, was
performed on SDR observations for the period 2016 - 2021. During the
optimization, eight simple clear sky radiation models were tested
(namely, Daneshyar-Paltridge-Proctor, Kasten-Czeplak, Haurwitz,
Berger-Duffie, Adnot-Bourges-Campana-Gicquel, Robledo-Soler, Kasten and
Ineichen-Perez), with a wide range of factors. These models are
described in more details by \citet{Reno2012} and evaluated by
\citet{Reno2016}. We found, that Haurwitz's model, adjusted with the
factor \(a = 0.965\) yields one of the lowest root mean squared errors
(RMSE), while the procedure, manages to characterize the majority of the
data. Thus, our clear sky reference is derived by the Eq.~\ref{eq:ahau}.
\begin{equation}
\text{SDR}_\text{CSref} = a \times \text{SDR}_\text{Haurwitz} = 0.965 \times 1098 \times \cos(\theta) \times \exp \left( \frac{ - 0.057}{\cos(\theta)} \right) \label{eq:ahau}
\end{equation} where: \(\text{SDR}_\text{CSref}\) is the reference clear
sky SDR, in \(\text{W}\,\text{m}^{-2}\) and \(\theta\) is the solar
zenith angle (SZA).

The criteria that were used to identify whether a measurement was taken
under clear-sky conditions are presented below. A data point is flagged
as ``clear-sky'' if all criteria are satisfied, otherwise it is
considered to be ``cloud-sky''. Each criterion was applied for a running
window of \(11\) consecutive one-minute measurements, and the
characterization is assigned to the central value of the window. Each
parameter, was calculated both from the observations and the reference
clear sky model, for each comparison. The allowable range of variation
is defined by the model-derived value of the parameter multiplied by a
factor plus an offset. The factors and the offsets were determined
empirically, by manually inspecting each filter's performance on
selected days and adjusting them accordingly during an iterative
process. The criteria are:

@@

Due to the significant measurement uncertainty when the Sun is near the
horizon, we have excluded all measurements with SZA greater than
\(85^\circ\). Moreover, due to some obstructions around the site (hills
and buildings), we excluded data with azimuth angle between \(58^\circ\)
and \(120^\circ\) with SZA greater than \(78^\circ\). On the latter
instances, Sun is systematically not visible from the instrument's
location. To make the measurements comparable throughout the dataset, we
adjusted all one-minute radiometric values to the mean Sun - Earth
distance. Subsequently, we made all measurements relative to the Total
Solar Irradiance (TSI) at \(1\,\text{au}\), in order to compensate for
the Sun's intensity variability, using a time series of satellite TSI
observations. The TSI data we used are part of the ``NOAA Climate Data
Record of Total Solar Irradiance'' dataset \citep{Coddington2005}. The
initial daily values of this dataset were interpolated to match the time
step of our measurements. In the final dataset \(23.3\%\) of the
1-minute data were identified as under clear-sky conditions and \(43\%\)
as under cloud-sky conditions.

In order to investigate the SDR trends, we implemented an appropriate
aggregation scheme to the 1-minute data to derive a series in coarser
time-scales. To preserve the representativeness of the data we used the
following criteria: a) we accept only days with more than 50\% of the
daytime measurements present and valid, b) on the dataset of clear- and
cloudy-skies, we included only days with more than 60\% of the daytime
measurements identified as clear or cloudy respectively, c) monthly
values were computed from daily means only when at least 20 days were
available. To create the daily and monthly climatological means, we
averaged the data based on the day of year and calendar month,
respectively. For the seasonal means we averaged the mean daily values
in each season (Winter: December - February, Spring: March - May, etc.).
Finally, each data set was deseasonalized by subtracting the
corresponding climatological annual cycle (daily or monthly) from the
actual data. To estimate the SZA effect on the SDR trends, the
one-minute data were aggregated in \(1^\circ\) SZA bins, separately for
the morning and afternoon hours, and then were deseasonalized as
mentioned above.

@@

\caption{Anomalies (\%) of the daily all-sky SDR, relative to climatological values for 1993 - 2023. The black line shows the long term linear trend.}\label{fig:trendALL}
\end{figure}

>>>>

Due to the significant measurement uncertainty when the Sun is near the
horizon, we have excluded all measurements with SZA greater than
\(85^\circ\). Moreover, due to obstructions around the site (hills and
buildings) which block the direct irradiance, we excluded data with
azimuth angle in the range \(58^{\circ}\) - \(120^{\circ}\) and with SZA
greater than \(78^{\circ}\). To make the measurements comparable
throughout the dataset, we adjusted all one-minute data to the mean Sun
- Earth distance. Subsequently, we adjusted all measurements to the
Total Solar Irradiance (TSI) at \(1\,\text{au}\), in order to compensate
for the Sun's intensity variability, using a time series of satellite
TSI observations. The TSI data we used are part of the ``NOAA Climate
Data Record of Total Solar Irradiance'' dataset \citep{Coddington2005}.
The initial daily values of this dataset were interpolated to match the
time step of our measurements.

In order to estimate the effect of the sky conditions on the long term
variability of SDR, we created three datasets by characterizing each
one-minute measurement with a corresponding sky-condition flag (i.e.,
all-sky, clear-sky and cloudy-sky). To identify the clear-cases we used
a method proposed by \citet{Long2000} and by \citet{Reno2016}, which
were adapted and configured for the site of Thessaloniki. We note that
all methods have some subjectivity in the definition of clear or cloudy
sky cases. As a result, the details of the definition are site specific
and they rely on a combination of thresholds and comparisons with ideal
radiation models and statistical analysis of different signal metrics.
The CSid algorithm was calibrated with the main focus, to identify the
presence of clouds. Despite the fine-tuning of the procedure, in a few
marginal cases false positive or false negative results were identified
by manual inspection. However, due to their small number, they cannot
affect the final results of the study. For completeness, we provide
below a brief overview of the clear-sky identification algorithm (CSid),
along with the site specific thresholds.

@@

To calculate the reference clear-sky \(\text{SDR}_\text{CSref}\) we used
the \(\text{SDR}_\text{Haurwitz}\) derived by the radiation model of
\citet{Haurwitz1945}, adjusted for our site with a factor \(a\)
(Eq.~\ref{eq:ahau}), estimated through an iterative optimization
process, as described by \citet{Long2000} and \citet{Reno2016}. The
target of the optimization was the minimization of a function \(f(a)\)
(Eq.~\ref{eq:minf}) and was accomplished with the algorithmic function
``optimise'', which is an implementation based on the work of
\citet{Brent1973}, from the library ``stats'' of the R programming
language \citep{RCT2023}. \begin{equation}
f(a) = \frac{1}{n}\sum_{i=1}^{n} ( \text{SDR}_{\text{CSid},i} - a \times \text{SDR}_{\text{testCSref},i} )^2 \label{eq:minf}
\end{equation} where: \(n\) is the total number of daylight data,
\(\text{SDR}_{\text{CSid},i}\) are the data identified as clear sky by
CSid, \(a\) is a site-specific adjustment factor, and
\(\text{SDR}_{\text{testCSref},i}\) is the SDR derived by any of the
tested clear-sky radiation models.

The optimization and the selection of the clear sky reference model, was
performed on SDR observations for the period 2016 - 2021. During the
optimization, eight simple clear sky radiation models were tested
(namely, Daneshyar-Paltridge-Proctor, Kasten-Czeplak, Haurwitz,
Berger-Duffie, Adnot-Bourges-Campana-Gicquel, Robledo-Soler, Kasten and
Ineichen-Perez), with a wide range of factors. These models are
described in more details by \citet{Reno2012} and evaluated by
\citet{Reno2016}. We found, that Haurwitz's model, adjusted with the
factor \(a = 0.965\) yields one of the lowest root mean squared errors
(RMSE), while the procedure manages to characterize successfully the
majority of the data. Thus, our clear sky reference is derived by the
Eq.~\ref{eq:ahau}. \begin{equation}
\text{SDR}_\text{CSref} = a \times \text{SDR}_\text{Haurwitz} = 0.965 \times 1098 \times \cos(\theta) \times \exp \left( \frac{ - 0.057}{\cos(\theta)} \right) \label{eq:ahau}
\end{equation} where \(\theta\) is the solar zenith angle (SZA).

The criteria that were used to identify whether a measurement was taken
under clear-sky conditions are presented below. A data point is flagged
as ``clear-sky'' if all criteria are satisfied; otherwise it is
considered as ``cloud-sky''. Each criterion was applied for a running
window of \(11\) consecutive one-minute measurements, and the
characterization was assigned to the central datum of the window. Each
parameter, was calculated both from the observations and the reference
clear sky model, for each comparison. The allowable range of variation
is defined by the model-derived value of the parameter multiplied by a
factor plus an offset. The factors and the offsets were determined
empirically, by manually inspecting each filter's performance on
selected days and adjusting them accordingly during an iterative
process. The criteria are listed below, together with the range of
values within which the respective parameter should fall in order to
raise the clear-sky flag:

@@

In the final dataset \(23.3\%\) of the 1-minute data were identified as
under clear-sky conditions and \(43\%\) as under cloud-sky conditions.

\hypertarget{aggregation-of-data-and-statistical-approach}{%
\subsection{Aggregation of data and statistical
approach}\label{aggregation-of-data-and-statistical-approach}}

In order to investigate the SDR trends which are the main focus of the
study, we implemented an appropriate aggregation scheme to the
one-minute data to derive a series in coarser time-scales. To preserve
the representativeness of the data we used the following criteria: a) we
accept only days with more than 50\% of the daytime measurements present
and valid, b) on the dataset of clear- and cloudy-skies, we included
only days with more than 60\% of the daytime measurements identified as
clear or cloudy respectively, c) monthly values were computed from daily
means only when at least 20 days were available. To create the daily and
monthly climatological means, we averaged the data based on the day of
year and calendar month, respectively. For the seasonal means we
averaged the mean daily values in each season (Winter: December -
February, Spring: March - May, etc.). Finally, each data set was
deseasonalized by subtracting the corresponding climatological annual
cycle (daily or monthly) from the actual data. To estimate the SZA
effect on the SDR trends, the one-minute data were aggregated in
\(1^\circ\) SZA bins, separately for the morning and afternoon hours,
and then were deseasonalized as mentioned above.

@@

\caption{Anomalies (\%) of the daily all-sky SDR from the climatological mean for the period1993 - 2023. The black line is the long term linear trend.}\label{fig:trendALL}
\end{figure}
